---
title: 'Generative simulations of two-way admixture models'
output:
  html_document:
    df_print: paged
---

### Goal

Evaluate the relative performances of FA, PC projections, qpAdm and structure (snmf) in simulations with known truth.  


```{r}
library(tfa) 
library(LEA)
library(admixr)
```


### Simulating data

The random generator simulates (continuous) genotypes for two ancient ancestral populations, sampled at random times in the past, and a mixed population sampled in present-days. The present day population is identified by 'times = 1', and the samples from ancestral populations have equal sizes. The population-level admixture coefficient is the parameter 'alpha'. 

```{r}
## simulate hidden factors and genotypes in two-way admixture models
r_geno_admix <- function (n = 200, #number of samples
                          p = 10000, #number of SNPs
                          alpha = 0.5, #Admixture coefficient 
                          K = 1, # number of factors
                          times = NULL, # sample dates (increasing order, from the oldest sample to the most recent one = 1)
                          sigma = 0.1, # residual noise
                          lambda = 1e-3, # drift parameter  
                          sd.U = 1.0, # SD for scores (vector of size K)
                          sd.V = 1.0 # SD for loadings
                          )
{
  
  if (length(times) != n) stop("Object times not of length n")
  
  # Covariance model
  C <- matrix(NA, n, n)
  for (i in 1:n){
    for (j in 1:n)  C[i,j] <- min(times[i], times[j])}
  
   # Building latent factors U (n x K)
  Sigma <- diag(x = sd.U^2, nrow = K, ncol = K)
  U <- MASS::mvrnorm(n, mu = rep(0, K), Sigma = Sigma)

  U[times == 1, 1] <- rnorm(sum(times == 1), 0, 1)
  ntm1 <- sum(times < 1)
  U1 <- c(rnorm(ntm1/2, (1-alpha)*sd.U[1], 1),
          rnorm(ntm1/2, -alpha*sd.U[1], 1))
  
  U[times < 1, 1] <- U1                     
  
  U[,1] <- scale(U[,1], scale = F)
  
  V <- MASS::mvrnorm(p, mu = rep(0, K), Sigma = sd.V^2 * diag(K))
  Epsilon <- apply(matrix(rep(0,p),nrow = 1), 2, function(x) rnorm(n,x,sigma))
  E_brownian <-  t(MASS::mvrnorm(p, mu = rep(0, n), Sigma = (1/lambda)*C))
  
  M =  U %*% t(V) + E_brownian + Epsilon
 
 #Returns: U = latent factors, V = loadings, M = data matrix 
  
  return(list(M = M,
              U = U,
              U1 = U1,
              V = V))
}
```

### Simulation example 

To simulate 200 individuals, 100 of which are from a present-day population resulting from an ancient admixture event between 2 ancestral populations. The admixture rates are 0.25 from population 1 and 0.75 from population  2. The parameter'sdu_1 = 5' indicates that the ancestral populations were moderately differentiated.

```{r}
sdu_1 <- 5
lambda_truth <- 15
sigma <- .1
sd_v <- sigma
alpha <- 0.25
```


```{r}
n = 200 #nb of individuals
tn <- sort(runif(n)) # sample dates = 1 - ages
tn[1:100] <- 1.5*tn[1:100] #past
tn[101:200] <- 1  #100 moderns with date = 1

simu <- r_geno_admix(n = n, 
                     p = 20000, 
                     alpha = alpha,
                     K = 3, 
                     times = tn, 
                     sd.U = c(sdu_1, 1.2, 1), 
                     lambda = lambda_truth, 
                     sd.V = sd_v, #.2
                     sigma = sigma)
```


The next plot shows the first two factors, corresponding to the ground truth to be recovered for factor and principal component methods.


```{r}
# ground truth
pop <- rep(1,n)
pop[simu$U[,1] < 0 & tn < 1] <- 0
pop[simu$U[,1] > 0 & tn < 1] <- 2
col <- NULL
col[pop == 1] <- "grey80" # present day population
col[pop == 0] <- "salmon2" # ancestral population
col[pop == 2] <- "darkblue" # ancestral population

plot(simu$U, cex = 0.5, col = col, pch = 19, main = "Ground truth")
```
```{r}
kmc <- c(mean(simu$U[pop == 0,1]), mean(simu$U[pop == 1,1]), mean(simu$U[pop == 2,1]))
adm <- (max(kmc) - kmc[kmc < max(kmc) & kmc > min(kmc)])/(max(kmc) - min(kmc))
cat("Estimated admixture from the ground truth: ")
a_GT <- min(1 - adm, adm)
a_GT
```

By using a probit link function, the continuous data are transformed into binary genotypes as follows: 

```{r}
Y <- simu$M
ages <- 1 - tn
# conversion as binary genotype (0 or 1, haploid data)
genotype <- rbinom(length(Y), 1, prob = as.numeric(pnorm(Y)))
Y <- matrix(genotype, nrow = dim(Y)[1])

F0 <- apply(Y[pop == 0, ], 2, mean)
F2 <- apply(Y[pop == 2, ], 2, mean)

fst = function(f0,f2){ Hs <- (f0*(1-f0) + f2*(1-f2))/2
H <- (f0 + f2)*(2 - f0 - f2)/4
1 - Hs/H}

FST <- fst(F0,F2)
cat("Nei's Fst: ")
print(mean(FST))
```

### Choosing lambda for FA

```{r}
fa  <- tfa(ages,
            Y,
            k = 2,
            lambda = 5e-2,
            center = TRUE)

r_2 <- choose_lambda(fa,
                     Y,
                     min_range = -5,
                     max_range = 0.01)
abline(v=log10(1e-3), col = "orange", lty = 2)
```



### Running FA analysis

According to the previous analysis, we run FA with `lambda = 5e-3`.

```{r}
library(tfa)
fa <- tfa(sample_ages = ages, 
          k = 2, 
          Y = Y,
          lambda = 5e-3)
Un <- fa$u
```


The estimated admixture coefficients are computed as follows:

```{r}
cat("Admixture coefficients from the tfa package:\n")
write.table(file = "./popc.txt",
            data.frame(paste("id", 1:100, sep = ""),
                       paste("pop", pop, sep = "")))
pop.c <- read.table("./popc.txt", as.is = TRUE)
colnames(pop.c) <- c("Instance.ID","Group.ID")
meta <- as.tfa_metadata(pop.c)

a_TFA <-  ancestry_coefficients(model = fa,
                      metadata = meta,
                      source = c("pop0", "pop2"), 
                      target = "pop1", individual = FALSE)
a_TFA
a_TFA <- min(a_TFA)
```

The graphical result is close to the ground truth. 

```{r}
plot(Un, col = col, 
     pch =  c(rep(19,100), rep(8, 100)), 
     cex = .6, 
     main = "FA", 
     xlab = "Factor 1", 
     ylab = "Factor 2",
     las = 1)
abline(0,0, col = "yellow4", lty = 2)
```


### PC projections of ancient samples on present-day samples 

To compute PC projections of ancient samples on present-day samples,  we start by performing a PCA on the present-day samples.

```{r}
pc <- prcomp(Y[pop == 1,], scale = FALSE) 

c_mean <- apply(Y[pop == 1,], 2, mean)
G <- t(t(Y) - c_mean)
```

The projections of ancient samples on present-day PCs are computed as follows.

```{r}
Vn <- rbind(G[pop != 1,] %*% pc$rotation , pc$x)
Vn <- prcomp(Vn[,1:2])$x
```

The estimated ancestry coefficient is close to the truth:

```{r}
kmc <- c(mean(Vn[pop == 0,1]), mean(Vn[pop == 1,1]), mean(Vn[pop == 2,1]))
adm <- (max(kmc) - kmc[kmc < max(kmc) & kmc > min(kmc)])/(max(kmc) - min(kmc))
cat("Coefficient d'admixture pc-proj: ")
a_PC <- min(1 - adm, adm)
a_PC
```
But the plot exhibits a shrinkage bias, in which ancient samples are mixed with present-day samples.

```{r}
plot(Vn, 
     main = "PCA",  
     xlab = "Axis 1", 
     ylab = "Axis 2",
     col = col, 
     cex = .6, 
     pch =  c(rep(19,100), rep(8, 100)),
     las = 1)

abline(0,0, col = "yellow4", lty = 2)
points(cbind(kmc,rep(0,3)), cex = 2, col = "grey20", lwd = 3, pch = 3)
```


### Using qpAdm with idealized outgroups

In this section, we run qpAdm (having admixtools installed in the OS). Before running the following code chunks, it is necessary to create a folder named **snps**.

Three artificial outgroup/reference individuals are created, differing from the 200 samples at 400 private SNPs: "og1" and "og2" are the closest outgroups to the ancestral populations and "ogall" is the most distant outgroup. The modified genotypic matrix is Yo.

```{r}
NPRIV = 400
NPRIV1 = round(NPRIV/4,0)
NPRIV2 = round(NPRIV/4,0)
NPRIVO = round(NPRIV/2,0)

Yo <- cbind(Y, matrix(0, nrow = nrow(Y), ncol = NPRIV))
og1 <- Yo[ which(pop == 0)[1] ,]
og1[ncol(Y) + (1:NPRIV1)] <- sample(1:2, NPRIV1, replace= TRUE)


og2 <- Yo[which(pop == 2)[1],]
og2[ncol(Y) + NPRIV1 + (1:NPRIV2)] <- sample(1:2, NPRIV2, replace= TRUE)

ogall <- og2
ogall[ncol(Y) + NPRIV1 + NPRIV2 + (1:NPRIVO)] <- sample(1:2, NPRIVO, replace= TRUE)

Yo <- rbind(Yo, og1, og2, ogall)

write.geno(R = Yo, output.file = "./snps/snps.geno")
```

Next, we need to create some files specific to the `admixtools` format.

```{r}
indfile <- data.frame(paste("I", 1:nrow(Yo), sep = ""), 
                      sample(c("M","F"), nrow(Yo), rep = TRUE), 
                      c(paste("pop", pop, sep = ""), "og0", "og2", "og_all"))
write.table(file = "./snps/snps.ind", indfile, quote = F, row.names = F, col.names = F)
```

and:

```{r}
snpfile <- data.frame(paste("rs", 1:ncol(Yo), sep = ""), 
                      rep(1,ncol(Yo)),
                      round((1:ncol(Yo))/ncol(Yo), digit = 5),
                      1:ncol(Yo),
                      rep("C", ncol(Yo)),
                      rep("G", ncol(Yo)))
write.table(file = "./snps/snps.snp", snpfile, quote = F, row.names = F, col.names = F)
```


The result is as follows:
```{r}
snps <- eigenstrat("./snps/snps")

result <- qpAdm(
  target = "pop1",
  sources = c("pop2","pop0"),
  outgroups = c("og0", "og2", "og_all"),
  data = snps
)

result$proportions
```

In summary, the estimated admixture coefficient is equal to


```{r}
cat("qpAdm estimated admixture coefficient: ")
a_qpADM <-  as.numeric(result$proportions[1,2])
a_qpADM 
```


### STRUCTURE (snmf) analysis

In this section, we estimate the admixture coefficient with the program `snmf` which is similar to STRUCTURE or ADMIXTURE, but faster and available in R. 

```{r}
write.geno(R = Y, output.file = "./y.geno")
```


```{r}
k = 2
project.snmf = snmf("./y.geno",
                    K = k, 
                    entropy = TRUE, 
                    repetitions = 1,
                    alpha = 1000)
```
```{r}
my.colors <- c("darkblue", "salmon2","grey80")
qmatrix <- Q(project.snmf, K = k)
cat("STRUCTURE admixture coefficient: ")
adm <- mean(qmatrix[101:200, 1])
a_STR <- min(adm, 1 - adm)
a_STR
```

```{r}
barchart(project.snmf, K = k, sort.by.Q = FALSE, 
         col = my.colors, border = NA, space = 0,
         xlab = "Individuals", ylab = "Ancestry proportions", 
         main = "Ancestry matrix") -> bp

axis(1, at = 1:length(bp$order), 
     labels = bp$order, las = 3, cex.axis = .4)
```

```{r}
remove.snmfProject("y.snmfProject")
```


```{r}
round(data.frame(
  true_rate = alpha, 
  ground_truth = a_GT,
  FA = a_TFA,
  PC = a_PC,
  qpADM = a_qpADM,
  structure = a_STR), digit = 4)
```


